Predicting Ensemble Location and Width in Binaural Recordings of Music with Convolutional Neural Networks

### 2024-06-19 paper/main.tex
Update paper

\usepackage[english]{babel}
\usepackage{authblk}
\usepackage[a4paper,top=2cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{gensymb}
\usepackage{textcomp}
\usepackage[backend=biber, style=nature]{biblatex}
\usepackage{csquotes}
\usepackage[toc,page]{appendix}

@@

The advancement of sophisticated machine learning techniques, particularly deep learning networks, has prompted an intriguing exploration of the extent to which these tools can emulate the human auditory system without relying on advanced spatial audio feature engineering, traditionally employed in audio source localization tasks \cite{yang_deepear_2022, vera-diaz_towards_2018, pang_multitask_2019}. To investigate this, the study developed an audio localization method based on a multi-task convolutional neural network (CNN) model \cite{lecun_handwritten_1989, lecun_deep_2015}.

@@

In contrast to the aforementioned studies, the proposed method is not constrained by the number of sources. Furthermore, the approach is not limited to speech and has been applied to a wide range of musical datasets, including instruments and vocals. In contrast to previous studies that primarily focused on individual sources, the proposed method does not aim to separate them, but rather considers them as an ensemble, or in this case, a musical ensemble. This approach is similar to how the real musical ensembles are arranged on stage. To the authors' knowledge, this is one of the first methods to localize ensemble width (see \cite{antoniuk2023blind} for the previous ensemble-width-related study), and the first to localize both ensemble position and width simultaneously using a multi-task model.

@@

In contrast, black-box methods typically employ minimal feature engineering and heavily rely on modern machine learning techniques to extract features and make predictions. These methods implicitly and internally extract features, without disclosing their purposes or mimicking the human auditory system. This lack of transparency and predictability, combined with their reliance on deep neural networks with numerous learning parameters, necessitates the development and evaluation using very large datasets. These datasets often comprise thousands of examples (e.g., 6300 examples in the TIMIT corpus \cite{garofolo_john_s_timit_1993} used in \cite{yang_deepear_2022, benaroya_binaural_2018, wang_binaural_2020, vecchiotti19, ma_exploiting_2017, pang_multitask_2019, ma_robust_2018, may_robust_2015}) or custom corpora consisting of hundreds of thousands of recordings used in \cite{antoniuk2023blind, zielinski_automatic_2022, zielinski_spatial_2022, zielinski_comparison_2020}.

@@

The experiments conducted in this study involved 23,040 binaural recordings of music. The binaural recordings were synthesized using 192 multi-track publicly-available music recordings \cite{noauthor_mixing_nodate} and 30 HRTF databases (see \nameref{appendix:a_hrtf} for a detailed list of HRTF databases used in this study). The number of tracks in multi-track recording ranged from 5 to 62, with median of 9. For each pair of a multi-track recording and an HRTF database, four binaural recordings were synthesized with different random ensemble parameters, namely location $\theta$ and width $\omega$, as defined in Section \ref{subsec:ensemble_definition}. Both parameters were drawn from a uniform random distribution. Furthermore, the tracks of the input multi-track recordings were randomly assigned to sound source positions ($\theta_i$) to enhance the diversity of the final binaural corpora.

@@

The network topology used in this study was highly inspired by a AlexNet convolutional neural network \cite{krizhevsky_imagenet_2012}. Although AlexNet network was designed for image classification, the binaural recordings in this study were converted into magnitude spectrograms, as described in Section \ref{subsec:feature_extraction}. The network is consisting of a series of convolutional units followed by a series of classification units.

@@

@@

@@

>>>>

\usepackage[english]{babel}
\usepackage{authblk}
\usepackage[a4paper,top=2cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{gensymb}
\usepackage{textcomp}
\usepackage[backend=biber, style=nature]{biblatex}
\usepackage{csquotes}
\usepackage[toc,page]{appendix}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{multirow}
\usepackage{tabularx}

@@

The advancement of sophisticated machine learning techniques, particularly deep learning networks, has spurred an intriguing exploration into the extent to which these tools can emulate the human auditory system. This exploration proceeds without reliance on the advanced spatial audio feature engineering traditionally used in audio source localization, as evidenced by references \cite{yang_deepear_2022, vera-diaz_towards_2018, pang_multitask_2019}. While the application of convolutional neural networks (CNNs) \cite{lecun_handwritten_1989} to audio signals --- especially in conjunction with spectrograms \cite{thomas_analyzing_2014, espi_exploiting_2015} or other techniques \cite{abdel-hamid_applying_2012, sainath_deep_2013} --- is not novel, it has been successfully employed in prior studies. Building on these foundations, this study developed an audio localization method by creating a multi-task CNN model.

@@

In contrast to the aforementioned studies, the proposed method is not constrained by the number of sources. Furthermore, the approach is not limited to speech and has been applied to a wide range of musical datasets, including instruments and vocals. In contrast to previous studies that primarily focused on individual sources, the proposed method does not aim to separate them, but rather considers them as a group, or in this case, a musical ensemble. This approach is similar to how the real musical ensembles are arranged on stage, thereby emphasizing practical applications in live settings. To the authors' knowledge, this is one of the first methods to localize ensemble width (see \cite{antoniuk2023blind} for the previous ensemble-width-related study), and the first to localize both ensemble position and width simultaneously using a multi-task model.

@@

In contrast, black-box methods generally employ minimal feature engineering, relying instead on advanced machine learning techniques to autonomously extract features and make predictions. These methods, while effective, do not explicitly reveal the features they extract nor necessarily mimic the human auditory system. This opacity and the unpredictable nature of the outcomes, coupled with their dependency on deep neural networks that have extensive learning parameters, require the use of very large datasets for development and evaluation. Such datasets often include thousands of examples, as seen in the TIMIT corpus with 6300 examples \cite{garofolo_john_s_timit_1993} used in multiple studies \cite{yang_deepear_2022, benaroya_binaural_2018, wang_binaural_2020, vecchiotti19, ma_exploiting_2017, pang_multitask_2019, ma_robust_2018, may_robust_2015}, or custom corpora comprising hundreds of thousands of recordings \cite{antoniuk2023blind, zielinski_automatic_2022, zielinski_spatial_2022, zielinski_comparison_2020}. This poses a significant challenge in assembling a sufficiently large and diverse collection of labeled binaural recordings. However, this challenge can be addressed through the synthesis of binaural sounds, as demonstrated in various studies \cite{antoniuk2023blind, zielinski_automatic_2022, zielinski_spatial_2022, zielinski_comparison_2020,yang_deepear_2022, ma_robust_2018} and further elaborated upon in Section \ref{subsec:synthesis} of this paper.

@@

The experiments conducted in this study involved 23,040 binaural recordings of music. The binaural recordings were synthesized using 192 multi-track publicly-available music recordings \cite{noauthor_mixing_nodate} and 30 HRTF databases (see Table \ref{table:hrtfs} in \nameref{appendix:a_hrtf} for a detailed list of HRTF databases used in this study). The number of tracks in multi-track recording ranged from 5 to 62, with median of 9. For each pair of a multi-track recording and an HRTF database, four binaural recordings were synthesized with different random ensemble parameters, namely location $\theta$ and width $\omega$, as defined in Section \ref{subsec:ensemble_definition}. Both parameters were drawn from a uniform random distribution. Furthermore, the tracks of the input multi-track recordings were randomly assigned to sound source positions ($\theta_i$) to enhance the diversity of the final binaural corpora. Before the synthesis, the tracks were equalized to $-23$ LUFS, in accordance with the ITU-R BS.1770 recommendation \cite{noauthor_recommendation_nodate}.

@@

@@

The network topology employed in this study, depicted in Figure \ref{fig:architecture}, was significantly influenced by the AlexNet convolutional neural network \cite{krizhevsky_imagenet_2012}. Although AlexNet was originally designed for image classification, this study adapted it for the audio prediction task by converting binaural recordings into magnitude spectrograms, as described in Section \ref{subsec:feature_extraction}. This conversion allowed the spectrograms to be treated as images, enabling their use in a standard image-prediction-like task.

The network accepts two spectrograms as inputs, and its architecture consists of a series of convolutional units followed by classification units, culminating in two output heads for predicting ensemble width and location. The architecture was designed in a multi-task fashion, enabling a single network to predict both ensemble parameters simultaneously. This multi-task capability is achieved by connecting the output of the final classification unit to two separate outputs, one for each ensemble parameter.

Despite the availability of widely used techniques for addressing overfitting, such as the Dropout Layer \cite{srivastava_dropout_nodate}, and for accelerating training, such as Batch Normalization \cite{ioffe_batch_2015}, neither technique was employed in this study due to their observed ineffectiveness for the specific prediction task at hand. Instead, a Global Average Pooling Layer \cite{lin_network_2014} was utilized, known for its proven capabilities in reducing overfitting. In this particular task, the inclusion of this layer significantly reduced overfitting, lowering the final mean absolute error score by $0.83\degree$ (average across 10 trials) compared to configurations using a Flattening Layer.

@@

Monte-Carlo cross validation (without replacement).

@@

{\renewcommand{\arraystretch}{1.1}
\begin{table}[!h]
  \begin{center}
    \caption{\label{table:hrtfs}List of HRTF sets used to synthesize binaural audio excerpts}
    \begin{tabularx}{\linewidth}{
      |>{\hsize=0.3\hsize}X|
      >{\hsize=0.7\hsize}X|
      >{\raggedright\hsize=1.3\hsize}X|
      >{\raggedright\hsize=0.6\hsize}X|
      >{\raggedright\arraybackslash\hsize=2.1\hsize}X|
      >{\hsize=1\hsize}X|
      }
      \cline{1-6}
      \textbf{No.} & \textbf{Type} & \textbf{Head}                             & \textbf{Radius {[}m{]}} & \textbf{Source}                                                                                                                                             & \textbf{Acronym}                \\
      \cline{1-6}
      1.           & Human         & Human subject                             & 1.2                     & \multirow{2}{\hsize}{RWTH Aachen University \cite{braren_high-resolution_2020}}                                                                             & \multirow{2}{\hsize}{AACHEN}    \\
      \cline{1-4}
      2.           & Artificial    & GRAS 45BB-4 KEMAR                         & 1                       &                                                                                                                                                             &                                 \\
      \cline{1-6}
      3.           & Human         & Subject 2                                 & 1.2                     & \multirow{4}{\hsize}{Austrian Academy of Sciences \cite{noauthor_hrtf-database_nodate}}                                                                     & \multirow{4}{\hsize}{ARI}       \\
      \cline{1-4}
      4.           & Human         & Subject 4                                 & 1.2                     &                                                                                                                                                             &                                 \\
      \cline{1-4}
      5.           & Human         & Subject 10                                & 1.2                     &                                                                                                                                                             &                                 \\
      \cline{1-4}
      6.           & Artificial    & ARI Printed Head                          & 1.2                     &                                                                                                                                                             &                                 \\
      \cline{1-6}
      7.           & Human         & Subject 012                               & 1                       & \multirow{3}{\hsize}{CIPIC Interface Laboratory, University of California \cite{algazi_cipic_2001}}                                                         & \multirow{3}{\hsize}{CIPIC}     \\
      \cline{1-4}
      8.           & Human         & Subject 015                               & 1                       &                                                                                                                                                             &                                 \\
      \cline{1-4}
      9.           & Human         & Subject 020                               & 1                       &                                                                                                                                                             &                                 \\
      \cline{1-6}
      10.          & Artificial    & Neumann KU 100                            & 0.9                     & NASA (2007) \cite{andreopoulou_inter-laboratory_2015}                                                                                                       & \multirow{2}{\hsize}{CLUBFRITZ} \\
      \cline{1-6}
      11.          & Artificial    & Neumann KU 100                            & 1.5                     & Helsinki University of Technology (2009) \cite{andreopoulou_inter-laboratory_2015}                                                                          &                                 \\
      \cline{1-6}
      12.          & Artificial    & FABIAN                                    & 1.47                    & \multirow{4}{\hsize}{Technical University Berlin, Huawei Technologies, Munich Research Centre, Sennheiser Electronic \cite{brinkmann_cross-evaluated_2019}} & \multirow{4}{\hsize}{HUTUBS}    \\
      \cline{1-4}
      13.          & Human         & Subject pp2                               & 1.47                    &                                                                                                                                                             &                                 \\
      \cline{1-4}
      14.          & Human         & Subject pp3                               & 1.47                    &                                                                                                                                                             &                                 \\
                   &               &                                           &                         &                                                                                                                                                             &                                 \\
      \cline{1-6}
      15.          & Human         & Subject 1003                              & 1.95                    & \multirow{2}{\hsize}{IRCAM, AKG \cite{noauthor_listen_2023}}                                                                                                & \multirow{2}{\hsize}{LISTEN}    \\
      \cline{1-4}
      16.          & Human         & Subject 1002                              & 1.95                    &                                                                                                                                                             &                                 \\
      \cline{1-6}
      17.          & Artificial    & KEMAR DB-4004 (DB-061)                    & 1.4                     & \multirow{2}{\hsize}{MIT \cite{gardne_hrtf_1994}}                                                                                                           & \multirow{2}{\hsize}{MIT}       \\
      \cline{1-4}
      18.          & Artificial    & KEMAR DB-4004 (DB-065)                    & 1.4                     &                                                                                                                                                             &                                 \\
      \cline{1-6}
      19.          & Human         & Subject 001                               & 1.5                     & \multirow{3}{\hsize}{Tohoku University \cite{watanabe_dataset_2014}}                                                                                        & \multirow{3}{\hsize}{RIEC}      \\
      \cline{1-4}
      20.          & Human         & Subject 002                               & 1.5                     &                                                                                                                                                             &                                 \\
      \cline{1-4}
      21.          & Artificial    & Koken SAMRAI                              & 1.5                     &                                                                                                                                                             &                                 \\
      \cline{1-6}
      22.          & Artificial    & Neumann KU 100                            & 1.2                     & \multirow{3}{\hsize}{University of York \cite{armstrong_perceptual_2018}}                                                                                   & \multirow{3}{\hsize}{SADIE II}  \\
      \cline{1-4}
      23.          & Human         & Subject H3                                & 1.2                     &                                                                                                                                                             &                                 \\
      \cline{1-4}
      24.          & Human         & Subject H4                                & 1.2                     &                                                                                                                                                             &                                 \\
      \cline{1-6}
      25.          & Artificial    & KEMAR                                     & 1                       & South China University of Technology \cite{yu_near-field_2018}                                                                                              & SCUT                            \\
      \cline{1-6}
      26.          & Artificial    & Neumann KU 100                            & 1                       & TH Köln  \cite{porschmann_spherical_2017}                                                                                                                   & TH Köln                         \\
      \cline{1-6}
      27.          & Artificial    & FABIAN                                    & 1.7                     & \multirow{2}{\hsize}{TU Berlin \cite{brinkmann_high_2017, wierstorf_free_2011}}                                                                             & \multirow{2}{\hsize}{TU Berlin} \\
      \cline{1-4}
      28.          & Artificial    & GRAS 45BA KEMAR                           & 1                       &                                                                                                                                                             &                                 \\
      \cline{1-6}
      29.          & Artificial    & GRAS 45BB-4 KEMAR - subject A attachment  & 1                       & \multirow{3}{\hsize}{Aalborg University; University of Iceland \newline \cite{spagnol_viking_2019,spagnol_viking_2020}}                                     & \multirow{3}{\hsize}{VIKING}    \\
      \cline{1-4}
      30.          & Artificial    & GRAS 45BB-4 KEMAR - subject B attachments & 1                       &                                                                                                                                                             &                                 \\
                   &               &                                           &                         &                                                                                                                                                             &                                 \\
      \cline{1-6}
    \end{tabularx}
  \end{center}
\end{table}
